---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bp

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/johnschwenck/bp.svg?branch=master)](https://travis-ci.com/johnschwenck/bp)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/johnschwenck/bp?branch=master&svg=true)](https://ci.appveyor.com/project/johnschwenck/bp)
<!-- badges: end -->

## bp: Blood Pressure Analysis for R
Cardiovascular disease (CVD) is the leading cause of death worldwide with Hypertension, specifically, affecting over 1.1 billion people annually. The goal of the \code{bp} package is to provide a comprehensive toolbox for analyzing blood pressure data using a variety of statistical methods and metrics in an effort to bring more clarity to CVD.

The \code{bp} package includes two sample data sets:

- `hypnos_data`: a sample of a larger HYPNOS study containing ABPM data for multiple subjects using continuous monitoring devices
- `bp_jhs`: a single-subject data set from a [2019 pilot study](https://dataverse.harvard.edu/dataverse/r4r) containing non-ABPM data from a self-monitoring Omron Evolv device 

### Installation

You can install the released version of bp from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("bp")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("johnschwenck/bp")
```

For installation with vignettes:
```r
devtools::install_github("johnschwenck/bp", build_vignettes = TRUE)
```

### Intended Functionality
The `bp` package is designed to allow the user to initialize a processed dataframe by specifying any combination of the following variables present in the user-supplied data set (with the minimum requirement that `SBP` and `DBP` are included). The package will then utilize the processed dataframe to calculate various metrics from medical and statistical literature. 

The package has the ability to make use of the following physiological variables (expressed as integers):

* Systolic Blood Pressure (`SBP`) measured in mmHg
* Diastolic Blood Pressure (`DBP`) measured in mmHg
* Heart Rate (`HR`) measured in bpm
* Pulse Pressure (`PP`) measured in mmHg which is calculated as SBP - DBP
* Mean Arterial Pressure (`MAP`) measured in mmHg
* Rate Pressure Product (`RPP`) which is calculated as SBP multiplied by resting HR

The data can be further refined on a more granular scale, depending on the type of data supplied. In most instances, ABPM data will include some kind of binary column corresponding to awake vs asleep which the user will assign when initializing the processed data. Further, most blood pressure data sets contain a timestamp associated with each reading. The additional variables are as follows:

* `DATE_TIME` combination such as `12/1/2020 13:42:07` (`as.POSIXct` format)
* `ID` of individuals, if more than one
* `VISIT` corresponding to the visit of each individual, if more than one (integer)
* `WAKE` as a binary indicator where 1 denotes awake and 0 denotes asleep (binary 1 or 0)

After all available variables are identified and processed, the resulting processed dataframe is used for all other functions. See examples below for further details.

### Available Metrics
The package will then utilize the above variables to calculate various metrics from medical and statistical literature in order to quantify and classify the variability of the readings into their respective categories of hypertension (normal, elevated, or hypertensive).

* `arv` - Average Real Variability
* `bp_mag` - Blood Pressure Magnitude (peak and trough)
* `bp_range` - Blood Pressure Range
* `cv` - Coefficient of Variation
* `sv` - Successive Variation
* `dip_calc` - Nocturnal Dipping \% and Classification

## Example - HYPNOS data

There are two main steps involved: 

1. Load and process data into a new usable dataframe for all further analysis using the `process_data` function 

```{r}
#devtools::install_github("johnschwenck/bp")
library(bp)

## Load hypnos_data
data(hypnos_data)

## Process hypnos_data
data <- process_data(hypnos_data, 
                     sbp = 'syst', 
                     dbp = 'diast', 
                     bp_datetime = 'date.time', 
                     hr = 'hr', 
                     pp = 'PP', 
                     map = 'MaP', 
                     rpp = 'Rpp', 
                     id = 'id', 
                     visit = 'Visit', 
                     wake = 'wake')

```
**NOTE:** the `process_data` function is insensitive to capitalization of the supplied data column names. For this example, even though the original column name "SYST" exists in the `hypnos_data`, "syst" is still an acceptable name to be given to the function as shown. For emphasis, all of the above column names were intentionally entered using the wrong capitalization.

`SBP` and `DBP` must be specified for any other functions to work properly.


2. Using the newly processed `data`, calculate various metrics. Now that the `hypnos_data` has been processed, all functions will now rely on `data` instead. The calculation of the nocturnal dipping classification is shown below:
```{r}
dip_calc(data)
```

